Comparing Hydrogen and Jet-A for an N+3 Turbofan with Water Recirculation using Gradient-Based Optimization

### 2023-05-10 conference-paper/main.tex

% ==================================================
%	Abstract
% ==================================================
\begin{abstract}
    % Advances in commercial propulsion technology are pushing the development of efficient high bypass ratio turbofan engines with larger overall pressure ratios and internal temperatures.
    Current trends suggest that geared ultra high bypass ratio turbofans are the next generation of commercial propulsion systems.
    These engines pose a difficult design problem due to high internal temperatures, large overall pressure ratios, and increasingly smaller cores.
    Furthermore, the emphasis on decreasing emissions has driven the exploration of hydrogen-powered aircraft, adding to the already challenging design space.
    Carrying and burning hydrogen introduces complexity and weight penalties that we must offset using the fuel's thermodynamic and chemical properties.
    In this study, we create a closed-loop water recirculation system with a zero-dimensional thermodynamic model and compare the benefits between Jet-A and hydrogen fuels.
    We perform a gradient-based optimization parameter sweep to explore the trade-offs between performance and emissions using both fuels with water recirculation.
    The results quantify the design space for next-generation propulsion concepts that can benefit from hydrogen fuel's advantageous properties to reduce emissions and improve performance.
    % The results quantify the design space for next-generation propulsion concepts that can take advantage of hydrogen fuel's advantageous thermodynamic properties to reduce emissions and improve performance.
\end{abstract}

@@

% NOTE: AL-PA, I turned this into a subfigure to save some space
\begin{figure}[hbt!]
    \centering
    \begin{subfigure}[t]{0.49\textwidth}
        \includegraphics[width=\textwidth]{injector.pdf}
        \caption{
            Water from the extractor is simply injected into the core flow upstream of the high-pressure compressor.
        }
        \label{fig:injector}
    \end{subfigure}
    \hspace{2pt}
    \begin{subfigure}[t]{0.49\textwidth}
        \includegraphics[width=\textwidth]{extractor.pdf}
        \caption{
            We extract a fraction of the moles of water from the out flow of duct 5.
            The extracted water is routed back upstream to the injector and the rest is exhausted out the nozzle.
        }
        \label{fig:extractor}
    \end{subfigure}
    % [x] TODO: AL-PA, Might want to add a bit more to this caption if you think it's necessary.
    \caption{Injector (left) and extractor (right) diagrams that show the inputs and output streams of each component.
        The inputs and outputs show which flow values are used to compute the amount of recirculated water.}
    \label{fig:extract_inject}
\end{figure}

@@

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=1.0\textwidth]{N3_CLVR_cycle.pdf}
    \caption{
        Layout of the N+3 engine core with the closed-loop vapor recovery.
        Black arrows are flow connections, blue boxes are cycle elements.
        Flow stations are identified with bubble text.
    }
    \label{fig:n3_clvr}
\end{figure}

@@

\begin{equation}
    T_{4,\rm{ratio}} = \frac{T_{4,\rm{TOC}}}{T_{4,\rm{RTO}}}.
    \label{eq:t4_ratio}
\end{equation}

@@

\begin{equation}
    V_{\rm{ratio}} = \frac{V_{\rm{core,ideal}}C_{v,\rm{core}}}{V_{\rm{bypass,ideal}}C_{v,\rm{bypass}}},
    \label{eq:vratio}
\end{equation}
where $V_\text{ideal}$ is the ideal nozzle exit velocity and $C_v$ is the nozzle velocity coefficient that accounts for non-ideal effects.
The final design variable is the water recovery fraction at CRZ ($X_{\rm{H_2O,CRZ}}$).
% [x] TODO: AL-PA, optimizing for fuel burn subject to net thrust constraint...
% [x] TODO: AL-PA, this isn't the only reason we consider just the water recovery at cruise.  It's also the longest flight segment and contributes the most to long-term performance.
We only include the water recovery fraction at cruise because it has the largest impact on the fuel mass flow rate objective function.
Additionally, the CRZ operating condition is the longest flight segment and contributes the most to long-term engine performance.
The range for each design variable and the associated units are shown in Table~\ref{tab:dv_table}.
% We add the water recovery fraction only at the CRZ point as a design variable since we are optimizing $\dot{m}_{\rm{fuel,CRZ}}$.

@@

% [x] TODO: AL-PA, Define the T4 ratio with an equation and reference the equation in the table
% [x] TODO: AL-PA, I thought we were only doing recovery fractions at CRZ??
% [x] TODO: AL-PA, Make sure you reference the equations for the DVs/cons in the descriptions so the reader can connect them.
% [x] TODO: PA-PA, explain variable ranges
\begin{table}[hbt!]
    \centering
    \caption{
        Multipoint optimization problem definition.
        The objective function is the fuel mass flow rate at the CRZ condition.
        The constraints are the net thrust at TOC and an upper bound on engine diameter.
    }
    \small
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{r l l r l}
        \toprule
                        & Variable/Function              & Description                                                              & Units            & Quantity \\
        \hline
        minimize        & $\dot{m}_{\rm{fuel,CRZ}} $     & Fuel flow rate at CRZ                                                    & \unit{lbm/s}     & 1        \\
                        &                                &                                                                          &                  &          \\
        with respect to & $X_{\rm{H_2O,CRZ}}$            & Water recovery mass fraction at CRZ                                      & -                & 1        \\
                        & $\rm{PR}_{\rm{fan,TOC}}$       & TOC fan pressure ratio                                                   & -                & 1        \\
                        & $\rm{PR}_{\rm{LPC,TOC}}$       & TOC low-pressure compressor pressure ratio                               & -                & 1        \\
                        & $\rm{OPR}_{\rm{TOC}}$          & TOC overall pressure ratio                                               & -                & 1        \\
                        & $T_{4,\rm{RTO}}$               & RTO combustor temperature                                                & \unit{\degree R} & 1        \\
                        & $T_{4,\rm{ratio}}$             & TOC-to-RTO temperature ratio (Equation~\eqref{eq:t4_ratio})              & -                & 1        \\
                        & $V_{\rm{ratio,CRZ}}$           & Core-to-bypass nozzle velocity ratio at CRZ (Equation~\eqref{eq:vratio}) & -                & 1        \\
        \cline{3-5}
                        &                                & Total                                                                    &                  & 7        \\
                        &                                &                                                                          &                  &          \\
        subject to      & $F_{\rm{net,TOC}} \geq 5800.0$ & Target net thrust at TOC                                                 & \unit{lbf}       & 1        \\
                        & $D_{\rm{fan}} \leq 100$        & Maximum Fan Diameter                                                     & \unit{inch^2}    & 1        \\
        \cline{3-5}
                        &                                & Total                                                                    &                  & 2        \\
        \bottomrule
    \end{tabular}
    \label{tab:opt_problem}
\end{table}

@@

\begin{table}[hbt!]
    \centering
    \caption{Optimization results of the N+3 engine for four optimizations using Jet-A and hydrogen fuel with and without water recirculation.
    }
    \small
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{r l l l l l l}
                    & Variable/Function         & Jet-A Dry & Jet-A Wet & \ce{H2} Dry & \ce{H2} Wet & Units           \\
        \toprule
        objective   & $\dot{m}_{\rm{fuel,CRZ}}$ & 0.639     & 0.606     & 0.233       & 0.221       & \si{lbm/s}      \\
                    & $\rm{TSEC}_{\rm{CRZ}}$    & 8186.254  & 7766.338  & 8295.236    & 7859.635    & \si{lbm/hr/lbf} \\
        \hline
        variables   & $X_{\ce{H2O},\text{CRZ}}$ & -         & 0.3       & -           & 0.17        & -               \\
                    & $\rm{PR}_{fan,TOC}$       & 1.285     & 1.292     & 1.288       & 1.297       & -               \\
                    & $\rm{PR}_{LPC,TOC}$       & 4.0       & 4.0       & 4.0         & 4.0         & -               \\
                    & $\rm{OPR}_{\rm{TOC}}$     & 61.012    & 61.188    & 60.660      & 61.952      & -               \\
                    & $T_{4,\text{ratio}}$      & 0.913     & 0.911     & 0.918       & 0.915       & -               \\
                    & $T_{4,\text{RTO}}$        & 3220.574  & -         & 3204.071    & -           & $^\circ$R       \\
                    & $V_\text{ratio,CRZ}$      & 1.35      & 1.35      & 1.35        & 1.35        & -               \\
        \hline
        constraints & $F_\text{net,TOC}$        & 5800.0    & 5800.0    & 5800.0      & 5800.0      & lbf             \\
                    & $D_\text{fan}$            & 100.0     & 100.0     & 100.0       & 100.0       & $in^2$          \\
        \bottomrule
    \end{tabular}
    \label{tab:dv_opt}
\end{table}

@@

% NOTE: PA-PA/AL replace this with a parallel axis plot?
% [x] TODO: AL-PA, Yeah let's do the parallel axis plot for the basic thermodynamic parameters
% AL-PA, We can make another plot for the water fractions and the dpQp study.
% [ ] TODO: AL-PA, Can you make the y-axis range for T4_ratio smaller so we can see the differences a bit better.  Maybe like 0.85 - 0.95
% [ ] TODO: AL-PA, Same comment as above but for FPR, maybe 1.28-1.32
% [ ] TODO: AL-PA, If a variable has bounds, can we make the y-axis limits match the bounds?
\begin{figure}[hbt!]
    \centering
    \includegraphics[width=1.0\textwidth]{N3_parallel_coords.pdf}
    \caption{
        Optimization results of the N+3 engine with and without no water recovery using Jet-A and H2 as the fuel.
        The red lines show Jet-A fuel and the blue lines show hydrogen fuel.
        The solid lines show the engine without water recovery and the dashed lines show the engine with water recovery.
    }
    \label{fig:parallel_coords}
\end{figure}

@@

% Thrust splits
% [x] TODO: PA-PA: Add thrust splits for each opt
% [x] TODO: AL-PA: I think this should be showing gross thrust from the fan divided by gross thrust from the core.
%                  The idea here is to show percentages of thrust generated by fan/core and how that changes with water recovery.
% \begin{table}[hbt!]
%     \centering
%     \caption{Thrust breakdown for each flight operating condition.
%         The net thrust, gross thrust, and ram drag are shown for each optimization result.
%     }
%     \small
%     \renewcommand{\arraystretch}{1.2}
%     \begin{tabular}{r l l l l l}
%         Point & Thrust Fraction & Jet-A Dry & Jet-A Wet & H2 Dry  & H2 Wet  \\
%         \toprule
%         TOC   & Core            & 6.47\%    & 6.05\%    & 5.95\%  & 5.47\%  \\
%               & Bypass          & 93.53\%   & 93.95\%   & 94.05\% & 94.53\% \\
%         \hline
%         RTO   & Core            & 6.31\%    & 5.87\%    & 5.77\%  & 5.26\%  \\
%               & Bypass          & 93.69\%   & 94.13\%   & 94.23\% & 94.74\% \\
%         \hline
%         SLS   & Core            & 6.11\%    & 5.68\%    & 5.58\%  & 5.09\%  \\
%               & Bypass          & 93.89\%   & 94.32\%   & 94.42\% & 94.91\% \\
%         \hline
%         CRZ   & Core            & 6.02\%    & 5.42\%    & 5.53\%  & 4.71\%  \\
%               & Bypass          & 93.98\%   & 94.58\%   & 94.47\% & 95.29\% \\
%         \bottomrule
%     \end{tabular}
%     \label{tab:thrust}
% \end{table}

@@

        \caption{
            Pressure loss sweep of duct 5 which is just upstream of the core nozzle.
            The figure shows the optimized TSEC values (blue) for varying levels of pressure loss due to the condensation of water compared to the optimization problem without water recovery (red).
            This shows the feasible space available for designing a water condenser while still benefiting from water recovery.

@@

        % [ ] TODO: AL-PA, The statement about the upper bound is weird.  If it hit the upper bound each time, either say it was constant or remove that part.

@@

        \caption{
            Pressure loss sweep of duct 5 which is just upstream of the core nozzle.

>>>>

% ==================================================
%	Abstract
% ==================================================
\begin{abstract}
  Advances in commercial propulsion technology led to the development of efficient high bypass ratio turbofan engines with larger overall pressure ratios and internal temperatures.
  Current trends suggest that geared ultra high bypass ratio turbofans are the next generation of commercial propulsion systems.
  Furthermore, the emphasis on decreasing emissions has driven the exploration of hydrogen-powered aircraft, adding to the already challenging design space.
  Carrying and burning hydrogen introduces complexity and weight penalties that we must offset using the fuel's thermodynamic and chemical properties.
  In this study, we create a closed-loop water recirculation system with a zero-dimensional thermodynamic model and compare the benefits between Jet-A and hydrogen fuels.
  We perform a gradient-based optimization parameter sweep to explore the trade-offs between performance and emissions using both fuels with water recirculation.
  The results quantify the design space for next-generation propulsion concepts that can take advantage of hydrogen fuel's advantageous thermodynamic properties to reduce emissions and improve performance.
  % Advances in commercial propulsion technology are pushing the development of efficient high bypass ratio turbofan engines with larger overall pressure ratios and internal temperatures.
  Current trends suggest that geared ultra high bypass ratio turbofans are the next generation of commercial propulsion systems.
  These engines pose a difficult design problem due to high internal temperatures, large overall pressure ratios, and increasingly smaller cores.
  Furthermore, the emphasis on decreasing emissions has driven the exploration of hydrogen-powered aircraft, adding to the already challenging design space.
  Carrying and burning hydrogen introduces complexity and weight penalties that we must offset using the fuel's thermodynamic and chemical properties.
  In this study, we create a closed-loop water recirculation system with a zero-dimensional thermodynamic model and compare the benefits between Jet-A and hydrogen fuels.
  We perform a gradient-based optimization parameter sweep to explore the trade-offs between performance and emissions using both fuels with water recirculation.
  The results quantify the design space for next-generation propulsion concepts that can benefit from hydrogen fuel's advantageous properties to reduce emissions and improve performance.
  % The results quantify the design space for next-generation propulsion concepts that can take advantage of hydrogen fuel's advantageous thermodynamic properties to reduce emissions and improve performance.
\end{abstract}

@@

% NOTE: AL-PA, I turned this into a subfigure to save some space
\begin{figure}[hbt!]
  \centering
  \begin{subfigure}[t]{0.49\textwidth}
    \includegraphics[width=\textwidth]{injector.pdf}
    \caption{
      Water from the extractor is simply injected into the core flow upstream of the high-pressure compressor.
    }
    \label{fig:injector}
  \end{subfigure}
  \hspace{2pt}
  \begin{subfigure}[t]{0.49\textwidth}
    \includegraphics[width=\textwidth]{extractor.pdf}
    \caption{
      We extract a fraction of the moles of water from the out flow of duct 5.
      The extracted water is routed back upstream to the injector and the rest is exhausted out the nozzle.
    }
    \label{fig:extractor}
  \end{subfigure}
  % [x] TODO: AL-PA, Might want to add a bit more to this caption if you think it's necessary.
  \caption{Injector (left) and extractor(right) diagrams that show the inputs and output streams of each component.
    The inputs and outputs show which flow values are used to compute the amount of recirculated water.}
  \label{fig:extract_inject}
\end{figure}

@@

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=1.0\textwidth]{N3_CLVR_cycle.pdf}
  \caption{
    Simplified layout of the N+3 engine cycle with the closed-loop vapor recovery.
    Black arrows are flow connections, blue boxes are cycle elements.
    Flow stations are identified with bubble text.
  }
  \label{fig:n3_clvr}
\end{figure}

@@

\begin{equation}
  T_{4,\rm{ratio}} = \frac{T_{4,\rm{TOC}}}{T_{4,\rm{RTO}}}
  \label{t4ratio}
\end{equation}

@@

\begin{equation}
  V_{\rm{ratio}} = \frac{V_{\rm{core,ideal}}C_{v,\rm{core}}}{V_{\rm{bypass,ideal}}C_{v,\rm{bypass}}}
  \label{eq:vratio}
\end{equation}
where $V_\text{ideal}$ is the ideal nozzle exit velocity and $C_v$ is the nozzle velocity coefficient that accounts for non-ideal effects.
The final design variable is the water recovery fraction at CRZ ($X_{\rm{H_2O,CRZ}}$).
% [x] TODO: AL-PA, optimizing for fuel burn subject to net thrust constraint...
% [x] TODO: AL-PA, this isn't the only reason we consider just the water recovery at cruise.  It's also the longest flight segment and contributes the most to long-term performance.
We only include the water recovery fraction at cruise because it has the largest impact on the fuel mass flow rate objective function.
Additionally, the CRZ operating condition is the longest flight segment and contributes the most to long-term engine performance.
The range for each design variable and the associated units are shown in Table~\ref{tab:dv_table}.
% We add the water recovery fraction only at the CRZ point as a design variable since we are optimizing $\dot{m}_{\rm{fuel,CRZ}}$.

@@

% [x] TODO: AL-PA, Define the T4 ratio with an equation and reference the equation in the table
% [x] TODO: AL-PA, I thought we were only doing recovery fractions at CRZ??
% [x] TODO: AL-PA, Make sure you reference the equations for the DVs/cons in the descriptions so the reader can connect them.
% [x] TODO: PA-PA, explain variable ranges
\begin{table}[hbt!]
  \centering
  \caption{
    Multipoint optimization problem definition.
    The objective function is the thrust specific energy consumption at the cruise condition.
    The constraints are net thrust and engine diameter constraint at the design point, TOC.
  }
  \small
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{r l l r l}
    \toprule
                    & Variable/Function              & Description                                                              & Units          & Quantity \\
    \hline
    minimize        & $\dot{m}_{\rm{fuel,CRZ}} $     & Fuel flow rate at CRZ                                                    & \si{lbm/s}     & 1        \\
                    &                                &                                                                          &                &          \\
    with respect to & $X_{\rm{H_2O,CRZ}}$            & Water recovery mass fraction at CRZ                                      & -              & 1        \\
                    & $\rm{PR}_{\rm{fan,TOC}}$       & TOC fan pressure ratio                                                   & -              & 1        \\
                    & $\rm{PR}_{\rm{LPC,TOC}}$       & TOC low-pressure compressor pressure ratio                               & -              & 1        \\
                    & $\rm{OPR}_{\rm{TOC}}$          & TOC overall pressure ratio                                               & -              & 1        \\
                    & $T_{4,\rm{RTO}}$               & RTO combustor temperature                                                & \si{\degree R} & 1        \\
                    & $T_{4,\rm{ratio}}$             & TOC-to-RTO temperature ratio (Equation~\eqref{t4ratio})                  & -              & 1        \\
                    & $V_{\rm{ratio,CRZ}}$           & Core-to-bypass nozzle velocity ratio at CRZ (Equation~\eqref{eq:vratio}) & -              & 1        \\
    \cline{3-5}
                    &                                & Total                                                                    &                & 7        \\
                    &                                &                                                                          &                &          \\
    subject to      & $F_{\rm{net,TOC}} \geq 5800.0$ & Target net thrust at TOC                                                 & \si{lbf}       & 1        \\
                    & $D_{\rm{fan}} \leq 100$        & Maximum Fan Diameter                                                     & \si{inch^2}    & 1        \\
    \cline{3-5}
                    &                                & Total                                                                    &                & 2        \\
    \bottomrule
  \end{tabular}
  \label{tab:opt_problem}
\end{table}

@@

\begin{table}[hbt!]
  \centering
  \caption{Optimization results of the N+3 engine for four optimizations using Jet-A and hydrogen fuel with and without water recirculation.
  }
  \small
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{r l l l l l l}
                & Variable/Function         & Jet-A Dry & Jet-A Wet & \ce{H2} Dry & \ce{H2} Wet & Units           \\
    \toprule
    objective   & $\dot{m}_{\rm{fuel,CRZ}}$ & 0.639     & 0.606     & 0.233       & 0.221       & \si{lbm/s}      \\
                & $\rm{TSEC}_{\rm{CRZ}}$    & 8186.254  & 7766.338  & 8295.236    & 7859.635    & \si{lbm/hr/lbf} \\
    \hline
    variables   & $x_{H2O,CRZ}$             & -         & 0.3       & -           & 0.17        & -               \\
                & $\rm{PR}_{fan,TOC}$       & 1.285     & 1.292     & 1.288       & 1.297       & -               \\
                & $\rm{PR}_{LPC,TOC}$       & 4.0       & 4.0       & 4.0         & 4.0         & -               \\
                & $\rm{OPR}_{\rm{TOC}}$     & 61.012    & 61.188    & 60.660      & 61.952      & -               \\
                & $T_{4,ratio}$             & 0.913     & 0.911     & 0.918       & 0.915       & -               \\
                & $T_{4,RTO}$               & 3220.574  & -         & 3204.071    & -           & $^\circ$R       \\
                & $V_{ratio,CRZ}$           & 1.35      & 1.35      & 1.35        & 1.35        & -               \\
    \hline
    constraints & $F_{net,TOC}$             & 5800.0    & 5800.0    & 5800.0      & 5800.0      & lbf             \\
                & $D_{Fan}$                 & 100.0     & 100.0     & 100.0       & 100.0       & $in^2$          \\
    \bottomrule
  \end{tabular}
  \label{tab:dv_opt}
\end{table}

@@

% NOTE: PA-PA/AL replace this with a parallel axis plot?
% [x] TODO: AL-PA, Yeah let's do the parallel axis plot for the basic thermodynamic parameters
% AL-PA, We can make another plot for the water fractions and the dpQp study.
% [ ] TODO: AL-PA, Can you make the y-axis range for T4_ratio smaller so we can see the differences a bit better.  Maybe like 0.85 - 0.95
% [ ] TODO: AL-PA, Same comment as above but for FPR, maybe 1.28-1.32
% [ ] TODO: AL-PA, If a variable has bounds, can we make the y-axis limits match the bounds?
\begin{figure}[hbt!]
  \centering
  \includegraphics[width=1.0\textwidth]{N3_parallel_coords.pdf}
  \caption{Optimization results of the N+3 engine with and without no water recovery using Jet-A and H2 as the fuel.
    The red lines show Jet-A fuel and the blue lines show \ce{H2} fuel.
    The solid lines show the engine without water recovery and the dashed lines show the engine with water recovery.}
  \label{fig:parallel_coords}
\end{figure}

@@

% Thrust splits
% [x] TODO: PA-PA: Add thrust splits for each opt
% [x] TODO: AL-PA: I think this should be showing gross thrust from the fan divided by gross thrust from the core.
%                  The idea here is to show percentages of thrust generated by fan/core and how that changes with water recovery.
\begin{table}[hbt!]
  \centering
  \caption{Thrust breakdown for each flight operating condition.
    The net thrust, gross thrust, and ram drag are shown for each optimization result.
  }
  \small
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{r l l l l l}
    Point & Thrust Fraction & Jet-A Dry & Jet-A Wet & H2 Dry  & H2 Wet  \\
    \toprule
    TOC   & Core            & 6.47\%    & 6.05\%    & 5.95\%  & 5.47\%  \\
          & Bypass          & 93.53\%   & 93.95\%   & 94.05\% & 94.53\% \\
    \hline
    RTO   & Core            & 6.31\%    & 5.87\%    & 5.77\%  & 5.26\%  \\
          & Bypass          & 93.69\%   & 94.13\%   & 94.23\% & 94.74\% \\
    \hline
    SLS   & Core            & 6.11\%    & 5.68\%    & 5.58\%  & 5.09\%  \\
          & Bypass          & 93.89\%   & 94.32\%   & 94.42\% & 94.91\% \\
    \hline
    CRZ   & Core            & 6.02\%    & 5.42\%    & 5.53\%  & 4.71\%  \\
          & Bypass          & 93.98\%   & 94.58\%   & 94.47\% & 95.29\% \\
    \bottomrule
  \end{tabular}
  \label{tab:thrust}
\end{table}

@@

    \caption{Pressure loss sweep of Duct5 which is just upstream of the water vapor extractor component.
      The figure shows the optimized TSEC values for varying levels of pressure loss due to the condensation of water compared to the optimization problem without water recovery.
      This shows the working space available for designing a water condenser while still gaining the benefits of water recovery.}

@@

@@

    \caption{Pressure loss sweep of Duct5 which is just upstream of the water vapor extractor component.
